Density-dependent selection and the limits of relative fitness

### 2018-08-27 manuscript.tex
l

To illustrate this consistency requirement, suppose that two identical types (same $b$, $c$ and $d$) are present, the first with small density $l_1\ll 1$ and the second with large density $l_2 \gg 1$. In this case, uncontested territories make up a negligible fraction of $U$, the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$, and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$. For consistency, the approximate per-capita growth rates in \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

Suppose that we used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$. Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, $\sum_j \langle x_j\rangle_a = l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$. Thus, the rare type would be pathologically predicted to decline in frequency. This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type. To avoid this, the effective means must be taken in a way that ensures that the total propagule density $\langle X \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

For the focal type $i$, we use the conditional dispersal distributions $p$ to calculate the effective mean, 
\begin{equation}
\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}.
\end{equation}
However, for nonfocal types $j\neq i$, we explicitly evaluate the $X$-dependence of the conditional dispersal probabilities, assuming that $X$ has the same conditional Poisson distribution $P(X|X\geq 2)$ for both $\langle \rangle_r$ and $\langle \rangle_a$. Starting with $\langle \rangle_r$, we first imagine that $X$ is a fixed constant to obtain
\begin{align}
\sum_{x_j} p(x_j|x_i=1,X)x_j=\frac{l_j}{L-l_i}(X-1)
\label{eq:raremonster1}
\end{align}
The right hand side is obtained by observing that the sum is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total. Applying our assumption about the distribution of $X$ gives
\begin{align}
\langle x_j \rangle_r&=\frac{l_j}{L-l_i}\sum_{X=2}^{\infty} P(X|X\geq 2) (X-1)\nonumber\\
&=\frac{l_j}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}},
\label{eq:meanxjrare}
\end{align}
where the last line follows from $P(X|X\geq 2) =\frac{1}{{1-(1+L)e^{-L}}}P(X)$ and $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=L-1+e^{-L}$. Substituting Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
\begin{equation}
\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
\end{equation}
where $R_i$ is defined in Eq.~\eqref{eq:Dr}.

Turning now to $\langle \rangle_a$, the mean focal abundance is 
\begin{align}
\langle x_i \rangle_{a}&=\sum_{x_i} p(x_i|x_i\geq 2)x_i \nonumber\\
&=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
\end{align}
For nonfocal types $j\neq i$, we have analogously to Eq.~\eqref{eq:raremonster1},
\begin{align}
\sum_{\mathbf x_i}  p(\mathbf x_i|x_i\geq 2,X) x_j&= \sum_{\mathbf x_i}  p(\mathbf x_i|X_i=X-x_i) x_j\nonumber\\
&= \frac{l_j(X-x_i)}{L-l_i}.
\end{align}
Assuming again that $X$ has the conditional distribution $P(X|X\geq 2)$ yields 
\begin{align}
\langle x_j \rangle_a&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \langle x_i \rangle_a \right]\nonumber\\
&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
\end{align}
Combining these results with Eqs.~\eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
\begin{equation}
\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
\end{equation}
where $A_i$ is defined in Eq.~\eqref{eq:Da}.

@@

Having derived the approximation Eq.~\eqref{eq:master}, we now evaluate its domain of validity. Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values. Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} respectively. We now evaluate these conditions.

We first calculate $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$. We have $\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
\begin{align}
\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2(x_j)&=\langle x_j^2 \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2\nonumber\\
&=\frac{l_j^2 + l_j}{C}-\frac{l_j^2}{C^2}\nonumber \\
&=\left(1-\frac{1}{C}\right)\frac{l_j^2}{C}+\frac{l_j}{C},\label{eq:varr}
\end{align}
and
\begin{align}
\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(x_j,x_k)&=\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\langle x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\nonumber\\
&=\frac{1}{C}\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\frac{l_jl_k}{C^2}\nonumber\\
&=\left(1-\frac{1}{C}\right)\frac{l_j l_k}{C}\qquad\qquad j\neq k. \label{eq:covr}
\end{align} 
Note that $1-1/C$ is negative because $C<1$. Decomposing the variance in $\sum_{j\neq i} c_j x_j$,
\begin{equation}
\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
\end{equation}
we obtain 
\begin{equation}
\frac{\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-\frac{1}{C})\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
\end{equation}

Eq.~\eqref{eq:cvr} reveals two important points. First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation. In this case, the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities. The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}). Second, if some of the $c_j$ are much larger than the others, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large. Specifically, in the presence of a rare,  strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eq. \eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}. Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when there are large differences in $c$.

\begin{figure}
\centering
\includegraphics[scale=0.8]{approx_details.pdf}
\caption{\label{fig:approx_details} Comparison of our $\tilde{q}$,$\hat{q}$ approximation with simulations, and also with the naive $\tilde{p}$,$\hat{p}$ approximation, as a function of the relative $c$ difference between two types. Our approximation breaks down in the presence of large $c$ differences. The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approximation --- growth rates are not equal in the neutral case $c_1=c_2$. Simulation procedure is the same as in Fig.~\ref{fig:simcomp}, with $U=10^5$.}
\end{figure}

Turning now to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{_{p({\mathbf x}|x_i\geq 2)}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{p}}^2(x_j)$, where $\sigma_{\hat{p}}^2(x_j)=l_j$ for $j\neq i$. Here  
\begin{equation}
\sigma_{\hat{p}}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
\end{equation}
where $D= 1-(1+l_i)e^{-l_i}$, and 
\begin{equation}
\frac{\sigma_{\hat{p}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{p}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
\end{equation}

Similarly to Eq.~\eqref{eq:cvr}, the right hand side of Eq. \eqref{eq:cva} is small for both low and high nonfocal densities. Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximation in this case. Again, the approximation breaks down in the presence of a rare, strong competitor (Fig.~\ref{fig:approx_details}).

\section*{Appendix B: Total density in the Lotka-Volterra competition model}

>>>>

To illustrate this consistency requirement, suppose that two identical types (same $b$, $c$ and $d$) are present, the first with small density $l_1\ll 1$ and the second with large density $l_2 \gg 1$. In this case, uncontested territories make up a negligible fraction of $U$; the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$; and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$. For consistency, the approximate per-capita growth rates in \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

Suppose that we naively used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$. Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, $\sum_j \langle x_j\rangle_a \approx l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$. Thus, the rare type would be predicted to decline in frequency even though it has identical traits. This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type. To avoid this, the effective means must be taken in a way that ensures that the total propagule density $\langle X \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

For the focal type $i$, we use the conditional dispersal distributions $p$ to calculate the effective mean, 
\begin{equation}
\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}. \label{eq:effectivefocal}
\end{equation}
However, for nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$. Specifically, we assume that $X$ follows a  Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$.

Starting with Eq.~\eqref{eq:meanfieldr}, we only need to evaluate $\langle x_j \rangle_r$ since $\langle x_i \rangle_r=1$. To evaluate the $X$-dependence separately, we first hold $X$ fixed to obtain
\begin{align}
\sum_{x_j} p(x_j|x_i=1,X)x_j=\frac{l_j}{L-l_i}(X-1) \qquad j\neq i.
\label{eq:raremonster1}
\end{align}
The right hand side is obtained by observing that the sum on the left is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total. We then take the with respect to $P(X|X\geq 2)$ to give
\begin{align}
\langle x_j \rangle_r&=\frac{l_j}{L-l_i}\sum_{X=2}^{\infty} P(X|X\geq 2) (X-1)\nonumber\\
&=\frac{l_j}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}},
\label{eq:meanxjrare}
\end{align}
where the last line follows from $P(X|X\geq 2) =\frac{1}{{1-(1+L)e^{-L}}}P(X)$ and $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=L-1+e^{-L}$. Substituting Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
\begin{equation}
\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
\end{equation}
where $R_i$ is defined in Eq.~\eqref{eq:Dr}.

Turning now to Eq.~\eqref{eq:meanfielda}, from Eq.~\eqref{eq:effectivefocal} the mean focal abundance is 
\begin{align}
\langle x_i \rangle_{a}&=\sum_{x_i} p(x_i|x_i\geq 2)x_i \nonumber\\
&=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
\end{align}
For nonfocal types $j\neq i$, we have analogously to Eq.~\eqref{eq:raremonster1},
\begin{align}
\sum_{\mathbf x_i}  p(\mathbf x_i|x_i\geq 2,X) x_j&= \sum_{\mathbf x_i}  p(\mathbf x_i|X_i=X-x_i) x_j\nonumber\\
&= \frac{l_j(X-x_i)}{L-l_i}.
\end{align}
Again taking the expection with respect to $P(X|X\geq 2)$ yields 
\begin{align}
\langle x_j \rangle_a&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \langle x_i \rangle_a \right]\nonumber\\
&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
\end{align}
Combining these results with Eqs.~\eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
\begin{equation}
\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
\end{equation}
where $A_i$ is defined in Eq.~\eqref{eq:Da}.

@@

Having derived the approximation Eq.~\eqref{eq:master}, we now evaluate its domain of validity. Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values. Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}.

We first consider the means and standard deviations in Eq.~\eqref{eq:meanfieldr}. We have $\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
\begin{align}
\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2(x_j)&=\langle x_j^2 \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2\nonumber\\
&=\frac{l_j^2 + l_j}{C}-\frac{l_j^2}{C^2}\nonumber \\
&=\left(1-\frac{1}{C}\right)\frac{l_j^2}{C}+\frac{l_j}{C},\label{eq:varr}
\end{align}
and
\begin{align}
\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(x_j,x_k)&=\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\langle x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\nonumber\\
&=\frac{1}{C}\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\frac{l_jl_k}{C^2}\nonumber\\
&=\left(1-\frac{1}{C}\right)\frac{l_j l_k}{C}\qquad\qquad j\neq k. \label{eq:covr}
\end{align} 
Note that $1-1/C$ is negative because $C<1$. Decomposing the variance in $\sum_{j\neq i} c_j x_j$,
\begin{equation}
\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
\end{equation}
we obtain 
\begin{equation}
\frac{\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-\frac{1}{C})\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
\end{equation}
Eq.~\eqref{eq:cvr} shows that, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation. The right hand side of Eq.~\eqref{eq:cvr} is then approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities. The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}).

Turning to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{p({\mathbf x}|x_i\geq 2)}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_j)$. For nonfocal types ($j\neq i$) $\sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_j)=l_j$, whereas for the focal type we have  
\begin{equation}
\sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
\end{equation}
where $D= 1-(1+l_i)e^{-l_i}$, and 
\begin{equation}
\frac{\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{p}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
\end{equation}
Similarly to Eq.~\eqref{eq:cvr}, the right hand side of Eq. \eqref{eq:cva} is small for both low and high nonfocal densities provided that the $c_j$ have similar magnitudes. Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximation in this case (Fig.~\ref{fig:approx_details}).

In both Eqs.~\eqref{eq:cvr} and \eqref{eq:cva}, the standard deviation in $\sum_{j\neq i} c_j x_j$ can be large relative to its mean if some of the $c_j$ are much larger than the others. Specifically, in the presence of a rare, strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eqs.~\eqref{eq:cvr} and ~\eqref{eq:cva} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}. Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when there are large differences in $c$.

\begin{figure}
\centering
\includegraphics[scale=0.8]{approx_details.pdf}
\caption{\label{fig:approx_details} Comparison of our effective mean approximation Eq.~\eqref{eq:master} with simulations, and also with the naive $\langle \rangle_r=\langle\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\rangle_a=\langle\rangle_{p({\mathbf x}|x_i\geq 2)}$ approximation, as a function of the relative $c$ difference between two types. Eq.~\eqref{eq:master} breaks down in the presence of large $c$ differences. The inset shows the pathology of the naive approximation --- growth rates for rare and common types are not equal in the neutral case $c_1=c_2$. Simulation procedure is the same as in Fig.~\ref{fig:simcomp}, with $U=10^5$.}
\end{figure}

\section*{Appendix B: Total density under Lotka-Volterra competition}
